Internal models of type theory and internal higher categories for the Univalent Foundations

### 2014-03-26 article-models-ttmodels.tex
changes in notation, most notably the change from identity functions to identity terms

  are involved -- we will write $\mfam{f}^\Gamma$ and $\mtm{f}^\Gamma$ to make

@@

  The identity functions are coded by}

@@

  \tag{identity functions}

@@

\subsection{The compatibility rules}
We need to do several things in this section:
\begin{itemize}
\item postulate that ordinary extension, weakening, substitution and identity
functions are compatible with the model constructors.
\item postulate that each of the model constructors is compatible with ordinary
extension, weakening, substitution and identity functions. Actually, we want
that every term of this type theory is compatible with those. I.e.~every term
is a functor/morphism of models. Also the sections.
\item postulate that the model constructors are compatible with each other,
so that they come to model ordinary extension, weakening, substitution and
identity functions respectively.
\item Note that the rules for compatibility with extension are going to explain which
model $\ctxext{A}{P}$ is by telling what the families, the terms, extension,
weakening, substitution and identity functions are. Likewise the rules for
compatibility with weakening and substitution do this for their respective
cases.
\item The easiest set of compatibility rules comes with weakening. The compatibility
rules that deal with substitution are likely going to have to do with the
Yoneda lemma, which we should be able to implement at some point.
\item The model $\subst{\tfemp{A}}{\mtm{A}}$ is going to be initial in the category of
all models over $A$. That means: whenever $P$ is a family over $\subst{\tfemp{A}}
{\mtm{A}}$ there will be a term of $P$. This means in particular that
$\subst{\tfemp{\emptyf}}{\mtm{\emptyf}}$ is going to be $\mctx$.
\end{itemize}

@@

\subsubsection{Extension acts as the identity on families of families}
The family of which $\tfext{A}$ is a term, is unfolded as
\begin{equation*}
\unfold{\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
\end{equation*}
Because we have the judgmental equality $\ctxext{\Gamma}{{A}{\mfam{A}}}
\jdeq \ctxext{{\Gamma}{A}}{\mfam{A}}$, there is a term
\begin{equation*}
\jhom
  {{{\Gamma}{A}}{\mfam{A}}}
  {\mfam{\mfam{A}}}
  {\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
\end{equation*}
By the rule asserting that families of families are families of extensions, we
have $\mfam{\mfam{A}}\jdeq\jcomp{}{\tfext{A}}{\mfam{A}}$. By the rule that
families over weakenings are weakenings of families, we have the judgmental
equalities
\begin{equation*}
\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}
\jdeq 
  \unfold{\jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}}
\jdeq 
  \jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}
\end{equation*}
Therefore, we can compare the term $\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}$
to the identity function $\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]$. 
The compatibility
rule for the action of extension on families of families asserts that
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{{\Gamma}{A}}{\mfam{A}}}
      {\jcomp{}{\tfext{A}}{\mfam{A}}}
      {\jcomp{}{\tfext{A}}{\mfam{A}}}
      {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
      {\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]}
    }
\end{equation*}
\emph{(Note: this rule might be a consequence of an explanation of what $\mfam{{A}{P}}$
is in general, but at the moment I don't see how to do this)}

\subsubsection{Extension acts as the identity on terms of families of families}
We now investigate the nature of the morphism.
\begin{equation*}
\jfhom
  {{{\Gamma}{A}}{\mfam{A}}}
  {\mfam{\mfam{A}}}
  {\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}}
  {\mtm{\mfam{A}}}
  {\subst{\tfext{A}}{\mtm{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mtm{\tfext{A}}^{\ctxext{\Gamma}{A}}}
\end{equation*}
Note that we have the judgmental equality 
$\mtm{\mfam{A}}\jdeq\jcomp{}{\tfext{A}}{\mtm{A}}$. Likewise, we have the
judgmental equality $\subst{\tfext{A}}{\mtm{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}
\jdeq\jcomp{}{\tfext{A}}{\mtm{A}}$. Thirdly, we have the judgmental equality
$\mfam{\tfext{A}}^{\ctxext{\Gamma}{A}}\jdeq\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]$.
We may combine these three facts with \autoref{hom-over-id-is-hom} to see that
\begin{equation*}
\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\jcomp{}{\tfext{A}}}{\mfam{A}}}
  {\jcomp{}{\tfext{A}}{\mtm{A}}}
  {\jcomp{}{\tfext{A}}{\mtm{A}}}
  {\mtm{\tfext{A}}^{\ctxext{\Gamma}{A}}}
\end{equation*}
Now we see that we can require that $\mtm{\tfext{A}}^{\ctxext{\Gamma}{A}}$ is
the identity function on $\jcomp{}{\tfext{A}}{\mtm{A}}$:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
    {{{{\Gamma}{A}}{\mfam{A}}}{\jcomp{}{\tfext{A}}}{\mfam{A}}}
    {\jcomp{}{\tfext{A}}{\mtm{A}}}
    {\jcomp{}{\tfext{A}}{\mtm{A}}}
    {\mtm{\tfext{A}}^{\ctxext{\Gamma}{A}}}
    {\idfunc[\jcomp{}{\tfext{A}}{\mtm{A}}]}
    }
\end{equation*}

\subsubsection{Weakening followed by substitution is the identity}
The weakening morphism $\tfwk{A}^0$ gives a weakened morphism
\begin{equation*}
\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\tfwk{A}^0}}
\end{equation*}
which can be composed with the morphism $\tfsubst{A}^0$. We require that this
composition is the identity function on $\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
      {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
      {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
      {\jcomp{}{\ctxwk{\mtm{A}}{\tfwk{A}^0}}{\tfsubst{A}^0}}
      {\idfunc[\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}]}
    }
\end{equation*}
Likewise, the weakening morphism $\tfwk{A}^1$ gives a weakened morphism
\begin{equation*}
\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\tfwk{A}^0}}
  {\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{\tfwk{A}^1}}
\end{equation*}
which can be composed with the the morphism $\tfsubst{A}^1$. We require that
this composition is the identity function on 
$\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}$. Note that this composition gives
a morphism over $\jcomp{}{\ctxwk{\mtm{A}}{\tfwk{A}^0}}{\tfsubst{A}^0}$, which
is the identity function on $\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$; so in fact
it is an ordinary morphism in the context
$\ctxext{{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}{\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}$.
Thus the inference rule we require to become valid is:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}{\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}}
      {\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}}
      {\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}}
      {\jcomp{}{\ctxwk{\mtm{A}}{\tfwk{A}^1}}{\tfsubst{A}^1}}
      {\idfunc[\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}]}
    }
\end{equation*}

@@

\subsection{Compatibility properties of arbitrary terms}
In this subsection we will state the inference rules that assert that every
term $\jterm{{\Gamma}{A}}{P}{f}$ acts functorially. That means, every term
preserves the empty context, extension, weakening, substitution and the
identity functions. Moreover, there will be several inference rules
involving the behavior of $\mfam{f}$ and $\mtm{f}$.

>>>>

  are involved -- we will write $\mfam[\Gamma]{f}$ and $\mtm[\Gamma]{f}$ to make

@@

  The identity terms are coded by}

@@

  \tag{identity terms}

@@

\subsection{The compatibility rules}
We need to do several things in this section:
\begin{itemize}
\item postulate that ordinary extension, weakening, substitution and identity
functions are compatible with the model constructors.
\item postulate that each of the model constructors is compatible with ordinary
extension, weakening, substitution and identity terms. Actually, we want
that every term of this type theory is compatible with those. I.e.~every term
is a functor/morphism of models. Also the sections.
\item postulate that the model constructors are compatible with each other,
so that they come to model ordinary extension, weakening, substitution and
identity terms respectively.
\item Note that the rules for compatibility with extension are going to explain which
model $\ctxext{A}{P}$ is by telling what the families, the terms, extension,
weakening, substitution and identity terms are. Likewise the rules for
compatibility with weakening and substitution do this for their respective
cases.
\item The easiest set of compatibility rules comes with weakening. The compatibility
rules that deal with substitution are likely going to have to do with the
Yoneda lemma, which we should be able to implement at some point.
\item The model $\subst{\tfemp{A}}{\mtm{A}}$ is going to be initial in the category of
all models over $A$. That means: whenever $P$ is a family over $\subst{\tfemp{A}}
{\mtm{A}}$ there will be a term of $P$. This means in particular that
$\subst{\tfemp{\emptyf}}{\mtm{\emptyf}}$ is going to be $\mctx$.
\end{itemize}

@@

\subsubsection{Extension acts as the identity on families of families}
The family of which $\tfext{A}$ is a term, is unfolded as
\begin{equation*}
\unfold{\jhom{\Gamma}{{A}{\mfam{A}}}{A}{\tfext{A}}}
\end{equation*}
Because we have the judgmental equality $\ctxext{\Gamma}{{A}{\mfam{A}}}
\jdeq \ctxext{{\Gamma}{A}}{\mfam{A}}$, there is a term
\begin{equation*}
\jhom
  {{{\Gamma}{A}}{\mfam{A}}}
  {\mfam{\mfam{A}}}
  {\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mfam[\ctxext{\Gamma}{A}]{\tfext{A}}}
\end{equation*}
By the rule asserting that families of families are families of extensions, we
have $\mfam{\mfam{A}}\jdeq\jcomp{}{\tfext{A}}{\mfam{A}}$. By the rule that
families over weakenings are weakenings of families, we have the judgmental
equalities
\begin{equation*}
\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}
\jdeq 
  \unfold{\jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}}
\jdeq 
  \jcomp{{A}{P}}{\tfext{A}}{\mfam{A}}
\end{equation*}
Therefore, we can compare the term $\mfam[\ctxext{\Gamma}{A}]{\tfext{A}}$
to the identity term $\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]$. 
The compatibility
rule for the action of extension on families of families asserts that
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{{\Gamma}{A}}{\mfam{A}}}
      {\jcomp{}{\tfext{A}}{\mfam{A}}}
      {\jcomp{}{\tfext{A}}{\mfam{A}}}
      {\mfam[\ctxext{\Gamma}{A}]{\tfext{A}}}
      {\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]}
    }
\end{equation*}
\emph{(Note: this rule might be a consequence of an explanation of what $\mfam{{A}{P}}$
is in general, but at the moment I don't see how to do this)}

\subsubsection{Extension acts as the identity on terms of families of families}
We now investigate the nature of the morphism.
\begin{equation*}
\jfhom
  {{{\Gamma}{A}}{\mfam{A}}}
  {\mfam{\mfam{A}}}
  {\subst{\tfext{A}}{\mfam{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mfam[\ctxext{\Gamma}{A}]{\tfext{A}}}
  {\mtm{\mfam{A}}}
  {\subst{\tfext{A}}{\mtm{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}}
  {\mtm[\ctxext{\Gamma}{A}]{\tfext{A}}}
\end{equation*}
Note that we have the judgmental equality 
$\mtm{\mfam{A}}\jdeq\jcomp{}{\tfext{A}}{\mtm{A}}$. Likewise, we have the
judgmental equality $\subst{\tfext{A}}{\mtm{\ctxwk{\ctxext{A}{\mfam{A}}}{A}}}
\jdeq\jcomp{}{\tfext{A}}{\mtm{A}}$. Thirdly, we have the judgmental equality
$\mfam[\ctxext{\Gamma}{A}]{\tfext{A}}\jdeq\idfunc[\jcomp{}{\tfext{A}}{\mfam{A}}]$.
We may combine these three facts with \autoref{hom-over-id-is-hom} to see that
\begin{equation*}
\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\jcomp{}{\tfext{A}}}{\mfam{A}}}
  {\jcomp{}{\tfext{A}}{\mtm{A}}}
  {\jcomp{}{\tfext{A}}{\mtm{A}}}
  {\mtm[\ctxext{\Gamma}{A}]{\tfext{A}}}
\end{equation*}
Now we see that we can require that $\mtm[\ctxext{\Gamma}{A}]{\tfext{A}}$ is
the identity term on $\jcomp{}{\tfext{A}}{\mtm{A}}$:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
    {{{{\Gamma}{A}}{\mfam{A}}}{\jcomp{}{\tfext{A}}}{\mfam{A}}}
    {\jcomp{}{\tfext{A}}{\mtm{A}}}
    {\jcomp{}{\tfext{A}}{\mtm{A}}}
    {\mtm[\ctxext{\Gamma}{A}]{\tfext{A}}}
    {\idfunc[\jcomp{}{\tfext{A}}{\mtm{A}}]}
    }
\end{equation*}

\subsubsection{Weakening followed by substitution is the identity}
The weakening morphism $\tfwk{A}^0$ gives a weakened morphism
\begin{equation*}
\jhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\tfwk{A}^0}}
\end{equation*}
which can be composed with the morphism $\tfsubst{A}^0$. We require that this
composition is the identity term on $\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
      {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
      {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
      {\jcomp{}{\ctxwk{\mtm{A}}{\tfwk{A}^0}}{\tfsubst{A}^0}}
      {\idfunc[\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}]}
    }
\end{equation*}
Likewise, the weakening morphism $\tfwk{A}^1$ gives a weakened morphism
\begin{equation*}
\jfhom
  {{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}
  {\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\mfam{\mfam{A}}}}
  {\ctxwk{\mtm{A}}{\tfwk{A}^0}}
  {\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}}
  {\ctxwk{\mtm{A}}{\mtm{\mfam{\mfam{A}}}}}
  {\ctxwk{\mtm{A}}{\tfwk{A}^1}}
\end{equation*}
which can be composed with the the morphism $\tfsubst{A}^1$. We require that
this composition is the identity term on 
$\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}$. Note that this composition gives
a morphism over $\jcomp{}{\ctxwk{\mtm{A}}{\tfwk{A}^0}}{\tfsubst{A}^0}$, which
is the identity term on $\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}$; so in fact
it is an ordinary morphism in the context
$\ctxext{{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}{\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}$.
Thus the inference rule we require to become valid is:
\begin{equation*}
\inference
  { \jfam{\Gamma}{A}
    }
  { \jhomeq
      {{{{{\Gamma}{A}}{\mfam{A}}}{\mtm{A}}}{\ctxwk{\mtm{A}}{{\mfam{A}}{\mfam{A}}}}}
      {\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}}
      {\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}}
      {\jcomp{}{\ctxwk{\mtm{A}}{\tfwk{A}^1}}{\tfsubst{A}^1}}
      {\idfunc[\ctxwk{\mtm{A}}{\ctxwk{\mfam{A}}{\mtm{A}}}]}
    }
\end{equation*}

@@

\subsection{Compatibility properties of arbitrary terms}
In this subsection we will state the inference rules that assert that every
term $\jterm{{\Gamma}{A}}{P}{f}$ acts functorially. That means, every term
preserves the empty context, extension, weakening, substitution and the
identity terms. Moreover, there will be several inference rules
involving the behavior of $\mfam{f}$ and $\mtm{f}$.
